An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-12-27 article/article.tex
finished chapter 2

Simulator-based models are particularly captivating due to the
provided modeling freedom. In essence, any data generating mechanism
that can be written as a finite set of algorithmic steps can be
programmed as a simulator-based model. In these cases, it is feasible
to generate samples using the simulator but it is infeasible to
evaluate the likelihood function. The intractability of the likelihood
makes the likelihood-free inference (LFI), i.e., the approximation of
the posterior distribution without using the likelihood function,
particularly challenging.

Optimization Monte Carlo (OMC) proposed by~\citet{Meeds2015} is a
novel LFI approach for approximating the posterior distribution. The
central idea is turning the stochastic data-generating mechanism into
a set of deterministic optimization processes. Afterwards,
\citet{Forneron2016} provided a similar method under the name `reverse
sampler'. In their work,~\citet{Ikonomov2019}, located some critical
limitations of OMC, so they proposed Robust OMC (ROMC), an alternative
version of OMC with the appropriate modifications.

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (Engine for likelihood-free
inference)~\citet{1708.00707}. As we illustrate at
Section~\ref{sec:implementation}, we have carefully designed the
implementation to ensure extensibility. ROMC is a general framework
for obtaining weighted samples from the posterior, i.e., it defines a
sequence of algorithmic steps without enforcing a specific algorithm
for solving each step. Therefore, a researcher may use ROMC as the
backbone algorithm and develop novel methods to solve each separate
step.\footnote{For being a ready-to-use
algorithm,~\citet{Ikonomov2019} proposed a default method for each
step, but this choice is by no means restrictive.} We have designed
our software for facilitating such experimentation.

To the best of our knowledge, this is the first implementation of the
ROMC inference method to a generic LFI framework. We organize the
evaluation in three steps. First, for securing that our implementation
is accurate, we test it against an artificial example with a tractable
likelihood. The artificial example also serves as a step-by-step guide
for showcasing how to use the various functionalities of our
implementation. Second, we use the second-order moving average (MA2)
example from the \pkg{ELFI} package, using as ground truth the samples
obtained with Rejection ABC~\citet{lintusaari2017}, with a very large
number of trials. Finally, we present the execution times of ROMC,
measuring the speed-up obtained by using the parallel version of the
implementation.

@@

We first give a short introduction to simulator-based models, we then
focus on OMC and its robust improvement, ROMC, and we, finally,
introduce \pkg{ELFI}, the underlying package that is used for the
implementation of ROMC.

@@

An implicit or simulator-based model is a parameterized stochastic
data generating mechanism. The key characteristic of simulator-based
models is that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterized
family of probability density functions
\(\{ p(\yb|\thetab)\}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these scenarios, we can only a
access the simulator \( m_r(\thetab) \), i.e., the black-box mechanism
(computer code) that generates samples \(\yb\) in a stochastic manner
from a set of parameters \(\thetab\). We denote the process of
obtaining samples from the simulator with
\( m_r(\thetab) \rightarrow \yb \). As shown by~\citet{Meeds2015}, it
is feasible to isolate the randomness of the simulator by introducing
a set of nuisance random variables denoted by \(\ub \sim
p(\ub)\). Therefore, for a specific tuple \((\thetab, \ub)\) the
simulator becomes a deterministic mapping \(g\), such that
\(\yb=\simulator(\thetab,\ub)\). In terms of computer code, the
randomness of a random process is governed by the global
seed. Although each software may handle the randomness in slightly
different ways\footnote{For example, at
  \pkg{Numpy}~\citet{harris2020array}, the pseudo-random number
  generation is based on a global state, whereas, in
  \pkg{JAX}~\citet{jax2018github} random functions consume a key that
  is passed as parameter. }, in all cases, freezing the initial seed
to a specific integer converts the simulation to a deterministic piece
of code.

Simulator-based models provide modeling freedom, including hidden
(unobserved) random variables or rule-based decisions. In fact, any
physical process that can be conceptualized as a computer program of
finite steps can be modeled as a simulator-based model. Hence,
implicit models are often used to model physical phenomena in the
natural sciences such as, e.g., genetics, epidemiology or
neuroscience. Further background on simulator-based models and example
applications can be found in the articles by~\citet{gutmann2016,
  lintusaari2017, sisson2018, cranmer2020}.

The modeling freedom of simulator-based models, however, comes at the
price of difficulties in inferring the parameters of
interest. Denoting the observed data as \(\data\), the main difficulty
lies at the intractability of the likelihood function
\(L(\thetab) = p(\data|\thetab)\). To better see the sources of the
intractability, and to address them, we go back to the basic
characterization of the likelihood as the (rescaled) probability of
generating data \(\yb\) that is similar to the observed data
\(\data\), using parameters \(\thetab\). Formally, the likelihood
\(L(\thetab)\) equals

\begin{equation}
  \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around
\(\data\) that is defined through a distance function \(d\), i.e., \
\(\region(\data) := \{\yb: d(\yb, \data) \leq \epsilon \}\).

Equation~\ref{eq:likelihood} highlights the main source of
intractability; computing
\(\Pr(\simulator(\thetab,\ub) \in \region(\data)) | \thetab\) as the
fraction of samples that lie inside the \(\epsilon\) region around
\(\data\) is computationally infeasible in the limit of
\(\epsilon \to 0\). Hence, the constraint is relaxed to
\(\epsilon > 0\), which leads to the approximate likelihood:

@@

\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data)
  \propto L_{d, \epsilon}(\thetab) p(\thetab)
  \propto L_{d, \epsilon}(\thetab) p(\thetab)
\end{equation}

@@

In Equation~\ref{eq:samples_approx_likelihood_omc}, for each
\(\ub_i\), there is a region \(\accregioni\) in the parameter space
\(\thetab\) where the indicator function returns one, i.e.,
\(\accregioni = \{ \thetab: \simulator(\thetab, \ub_i) \in
\region(\data) \}\). Therefore, we can approximate the likelihood as:

@@

and the corresponding approximate posterior:

@@

ROMC approximates the acceptance regions \(\accregioni\) with
\(D\)-dimensional bounding boxes \(\accregionihat\). A uniform
distribution, \(q_i(\thetab)\), is defined on top of each bounding
box. Each \(q_i(\thetab)\) serves a proposal distribution for
generating posterior samples \(\thetab_{ij} \sim q_i\). The samples
get an (importance) weight \(w_{ij}\) that compensate for using the
proposal distributions \(q_i(\thetab)\) instead of the prior
\(p(\thetab)\):

@@

Given the weighted samples, any expectation
\(\E_{p(\thetab|\data)}[h(\thetab)]\) of some function \(h(\thetab)\), can be approximated as
\begin{equation} \label{eq:expectation}
  \E_{p(\thetab|\data)}[h(\thetab)] \approx \frac{\sum_{ij} w_{ij} h(\thetab_{ij})}{\sum_{ij} w_{ij}}
\end{equation}
\citet{Ikonomov2019} considered uniform distributions as proposal
distributions so that the main task is to approximate the acceptance
regions \(\accregioni\) and to represent them so that uniform sampling
is easy. The approximation of the acceptance regions contains two
compulsory and one optional step: (i) solving the optimization
problems as in OMC, (ii) constructing bounding boxes around
\(\accregioni\) and, optionally, (iii) refining the approximation via a
surrogate model of the distance.

\subsubsection*{(i) Solving the deterministic optimization problems}
For each set of nuisance variables \(\ub_i, i = \{1,2,\ldots,n_1 \}\),
we search for a point \(\thetab^*_i\) such that
\(d(\simulator(\thetab^*_i,\ub_i), \data) \le \epsilon\). In
principle, \(d(\cdot)\) can refer to any valid distance function. For
the rest of the paper we consider \(d(\cdot)\) as the squared
Euclidean distance, as in~\citet{Ikonomov2019}. For easier notation,
we use \(d_i(\thetab)\) to refer to
\(d(\simulator(\thetab,\ub_i), \data)\). We search for \(\theta_i^*\)
solving the following optimization problem:

\begin{equation}
  \label{eq:optProb}
  \thetab_i^* = \!\argmin_{\thetab} d_i(\thetab)
\end{equation}
%
The solution is accepted only if it satisfies the constraint
\(d_i(\thetab_i^*) \le \epsilon\). If not, there is no point into
building an acceptance region around \(\accregioni\), so
\(\thetab_i^*\) is discarded. There are two proposed options for
solving Equation~\ref{eq:optProb} depending on the differentiability
of \(d_i(\thetab)\). If \(d_i(\thetab)\) is differentiable any
gradient-based optimizer can be used for
Equation~\ref{eq:optProb}. The gradients
\(\nabla_{\thetab} d_i(\thetab)\) can be either provided in closed
form or approximated by finite differences. If \(d_i\) is not
differentiable, Bayesian Optimization~\citep{Shahriari2016} can be
used instead. In this scenario, apart from obtaining an optimal
\(\thetab_i^* \), we can also automatically build a surrogate model
\(\hat{d}_i(\thetab)\) of the distance function \(d_i(\thetab)\). The
surrogate model \(\hat{d}_i\) can then substitute the actual distance
function in downstream steps of the algorithms, with possible
computational gains especially in cases where evaluating the actual
distance \(d_i(\thetab)\) is expensive.

@@

Each acceptance region \(\accregioni\) is approximated by a bounding
box \(\accregionihat\). The acceptance regions \(\accregioni\) can
contain any number of disjoint subsets in the \(D\)-dimensional space
and any of these subsets can take any arbitrary shape. First, since
the bounding boxes are built around \(\thetab_i^*\), we focus only on
the connected subset of \(\accregioni\) that contains \(\thetab_i^*\),
which we denote as \(\accregionitheta\). Second, we want to ensure
that the bounding box \(\accregionihat\) is big enough to contain on
its interior all the volume of \(\accregionitheta\). Third, we want to
be as tight as possible to \(\accregionihat\) to ensure high
acceptance rate on the importance sampling step that will
follow. Therefore, the bounding boxes are built as follows. First, we
compute their axes \(\mathbf{v}_m\), for \(m = \{1, \ldots, D\}\)
based on the (estimated) curvature of the distance at \(\thetab_i^*\),
and, second, we apply a line-search method along each axis to determine
the size of the box, see Algorithm~\ref{alg:region_construction} for
the details. After the bounding boxes construction, a uniform
distribution \(q_i\) is defined on each bounding box, and is used as
the proposal region for importance sampling.

@@

All inference methods are implemented as separate \pkg{ELFI} classes
with the two conventions:

\begin{itemize}
\item their constructor is \code{elfi.<Class name>(<output node>, *arg)};
  the initial argument is the output node of the simulator-based model
  and the rest of the arguments are parameters of the method.
\item they obtain posterior samples using \code{elfi.<Class name>.sample(*args)}
\end{itemize}

>>>>

Simulator-based models are particularly captivating due to the
modeling freedom they provide. In essence, any data generating
mechanism that can be written as a finite set of algorithmic steps can
be programmed as a simulator-based model. Hence, these models are
often used to model physical phenomena in the natural sciences such
as, e.g., genetics, epidemiology or neuroscience~\citet{gutmann2016,
  lintusaari2017, sisson2018, cranmer2020}. In simulator-based models,
it is feasible to generate samples using the simulator but is
infeasible to evaluate the likelihood function. The intractability of
the likelihood makes the so-called likelihood-free inference (LFI),
i.e., the approximation of the posterior distribution without using
the likelihood function, particularly challenging.

Optimization Monte Carlo (OMC), proposed by~\citet{Meeds2015}, is a
novel LFI approach. The central idea is to convert the stochastic
data-generating mechanism into a set of deterministic optimization
processes. Afterwards, \citet{Forneron2016} described a similar method
under the name `reverse sampler'. In their work,~\citet{Ikonomov2019},
identified some critical limitations of OMC and proposed Robust OMC
(ROMC) as an improved version of OMC with appropriate
modifications.

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (Engine for likelihood-free
inference)~\citet{1708.00707}. The implementation has been designed to
ensure extensibility. ROMC is an LFI framework; it defines a sequence
of algorithmic steps for aproximating the posterior without enforcing
a specific algorithm for solving each step. Therefore, a researcher
may use ROMC as the backbone algorithm and develop novel methods to
solve each separate step.\footnote{For being a ready-to-use
  algorithm,~\citet{Ikonomov2019} proposed a default method for each
  step, but this choice is by no means restrictive.} We have designed
our software for facilitating such experimentation.

To the best of our knowledge, this is the first implementation of the
ROMC inference method to a generic LFI framework. We organize the
evaluation in three steps. First, for securing that our implementation
is accurate, we test it against an artificial example with a tractable
likelihood. The artificial example also serves as a step-by-step guide
for showcasing how to use the various functionalities of our
implementation. Second, we use the second-order moving average (MA2)
example from the \pkg{ELFI} package, using as ground truth the samples
obtained with Rejection ABC~\citet{lintusaari2017} using a very high
number of samples. Finally, we present the execution times of ROMC,
measuring the speed-up achieved by the parallel version of the
implementation.

@@

We first give a short introduction to simulator-based models, we then
focus on OMC and its robust version, ROMC, and we, finally, introduce
\pkg{ELFI}, the \proglang{Python} package used for the implementation.

@@

An implicit or simulator-based model is a parameterized stochastic
data generating mechanism, where we can sample data points but we
cannot evaluate the likelihood. Formally, a simulator-based model is a
parameterized family of probability density functions
\(\{ p(\yb|\thetab)\}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these cases, we can only access the
simulator \( m_r(\thetab) \), i.e., the black-box mechanism (computer
code) that generates samples \(\yb\) in a stochastic manner from a set
of parameters \(\thetab\). We denote the process of obtaining samples
from the simulator with \( m_r(\thetab) \rightarrow \yb \). As shown
by~\citet{Meeds2015}, it is feasible to isolate the randomness of the
simulator by introducing a set of nuisance random variables denoted by
\(\ub \sim p(\ub)\). Therefore, for a specific tuple
\((\thetab, \ub)\) the simulator becomes a deterministic mapping
\(g\), such that \(\yb=\simulator(\thetab,\ub)\). In terms of computer
code, the randomness of a random process is governed by the global
seed. Although each package may handle the randomness in slightly
different ways\footnote{For example, at
  \pkg{Numpy}~\citet{harris2020array}, the pseudo-random number
  generation is based on a global state, whereas, in
  \pkg{JAX}~\citet{jax2018github} random functions consume a key that
  is passed as parameter. }, in all cases, setting the initial seed to
a specific integer converts the simulation to a deterministic piece of
code.

As stated at the Introduction, the modeling freedom of simulator-based
models comes at the price of difficulties in inferring the parameters
of interest. Denoting the observed data as \(\data\), the main
difficulty lies at the intractability of the likelihood function
\(L(\thetab) = p(\data|\thetab)\). To better see the sources of the
intractability, and to address them, we go back to the basic
characterization of the likelihood as the (rescaled) probability of
generating data \(\yb\) that is similar to the observed data
\(\data\), using parameters \(\thetab\). Formally, the likelihood
\(L(\thetab)\) is:

\begin{equation}
  \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around
\(\data\) that is defined through a distance function \(d\), i.e., \
\(\region(\data) := \{\yb: d(\yb, \data) \leq \epsilon
\}\). Equation~\ref{eq:likelihood} highlights the main source of
intractability; computing
\(\Pr(\simulator(\thetab,\ub) \in \region(\data) | \thetab) \) as the
fraction of samples that lie inside the \(\epsilon\) region around
\(\data\) is computationally infeasible in the limit where
\(\epsilon \to 0\). Hence, the constraint is relaxed to
\(\epsilon > 0\), which leads to the approximate likelihood:

@@

\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data)
  \propto L_{d, \epsilon}(\thetab) p(\thetab)
\end{equation}

@@

In Equation~\ref{eq:samples_approx_likelihood_omc}, for each
\(\ub_i\), there is a region \(\accregioni\) in the parameter space
\(\thetab\) where the indicator function returns one, i.e.,
\(\accregioni = \{ \thetab: \simulator(\thetab, \ub_i) \in
\region(\data) \}\). Therefore, we can rewrite the approximate
likelihood and posterior as:

@@

@@

ROMC approximates the acceptance regions \(\accregioni\) with
\(D\)-dimensional bounding boxes \(\accregionihat\). A uniform
distribution, \(q_i(\thetab)\), is defined on top of each bounding box
and serves as a proposal distribution for generating posterior samples
\(\thetab_{ij} \sim q_i\). The samples get an (importance) weight
\(w_{ij}\) that compensate for using the proposal distributions
\(q_i(\thetab)\) instead of the prior \(p(\thetab)\):

@@

Given the weighted samples, any expectation
\(\E_{p(\thetab|\data)}[h(\thetab)]\) of some function \(h(\thetab)\), can be approximated as
\begin{equation} \label{eq:expectation}
  \E_{p(\thetab|\data)}[h(\thetab)] \approx \frac{\sum_{ij} w_{ij} h(\thetab_{ij})}{\sum_{ij} w_{ij}}
\end{equation}

The approximation of the acceptance regions contains two
compulsory and one optional step: (i) solving the optimization
problems as in OMC, (ii) constructing bounding boxes around
\(\accregioni\) and, optionally, (iii) refining the approximation via a
surrogate model of the distance.

\subsubsection*{(i) Solving the deterministic optimization problems}
For each set of nuisance variables \(\ub_i, i = \{1,2,\ldots,n_1 \}\),
we search for a point \(\thetab^*_i\) such that
\(d(\simulator(\thetab^*_i,\ub_i), \data) \le \epsilon\). In
principle, \(d(\cdot)\) can refer to any valid distance function. For
the rest of the paper we consider \(d(\cdot)\) as the squared
Euclidean distance, as in~\citet{Ikonomov2019}. For simplicity, we use
\(d_i(\thetab)\) to refer to \(d(\simulator(\thetab,\ub_i),
\data)\). We search for \(\theta_i^*\) solving:

\begin{equation}
  \label{eq:optProb}
  \thetab_i^* = \!\argmin_{\thetab} d_i(\thetab)
\end{equation}
%
and we accept the solution only if it satisfies the constraint
\(d_i(\thetab_i^*) \le \epsilon\). If \(d_i(\thetab)\) is
differentiable, Equation~\ref{eq:optProb} can be solved using any
gradient-based optimizer. The gradients
\(\nabla_{\thetab} d_i(\thetab)\) can be either provided in closed
form or approximated by finite differences. If \(d_i\) is not
differentiable, Bayesian Optimization~\citep{Shahriari2016} can be
used instead. In this scenario, apart from obtaining an optimal
\(\thetab_i^* \), we can also automatically build a surrogate model
\(\hat{d}_i(\thetab)\) of the distance function \(d_i(\thetab)\). The
surrogate model \(\hat{d}_i\) can then substitute the actual distance
function in downstream steps of the algorithms, with possible
computational gains especially in cases where evaluating the actual
distance \(d_i(\thetab)\) is expensive.

@@

Each acceptance region \(\accregioni\) is approximated by a bounding
box \(\accregionihat\). The acceptance regions \(\accregioni\) can
contain any number of disjoint subsets in the \(D\)-dimensional space
and any of these subsets can take any arbitrary shape. We should make
three important remarks. First, since the bounding boxes are built
around \(\thetab_i^*\), we focus only on the connected subset of
\(\accregioni\) that contains \(\thetab_i^*\), which we denote as
\(\accregionitheta\). Second, we want to ensure that the bounding box
\(\accregionihat\) is big enough to contain on its interior all the
volume of \(\accregionitheta\). Third, we want \(\accregionihat\) to
be as tight as possible to \(\accregionitheta\) to ensure high
acceptance rate on the importance sampling step that
follows. Therefore, the bounding boxes are built in two steps. First,
we compute their axes \(\mathbf{v}_m\), for \(m = \{1, \ldots, D\}\)
based on the (estimated) curvature of the distance at \(\thetab_i^*\),
and, second, we apply a line-search method along each axis to
determine the size of the box. We refer the reader to
Algorithm~\ref{alg:region_construction} for the details. After the
bounding boxes construction, a uniform distribution \(q_i\) is defined
on each bounding box, and is used as the proposal region for
importance sampling.

@@

All inference methods of \pkg{ELFI} are implemented following two
conventions. First, their constructor follows the signature
\code{elfi.<Class name>(<output node>, *arg)}, where \code{<output
  node>} is the output node of the simulator-based model and
\code{*arg} are the parameters of the method. Second, they provide a
method \code{elfi.<Class name>.sample(*args)} for drawing samples from
the approximate posterior.
